Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming

### 2022-08-15 Paper/Manuscript_Stats&Comp/manuscript_STCO.tex
updated manuscript submitted

\institute{Gabriel Riutort-Mayol \at
              Foundation for the Promotion of Health and Biomedical Research of Valencia Region (FISABIO), Spain\\
              \emph{Corresponding Author} \\
              \email{gabriuma@gmail.com; riutort\_gab@gva.es}       %  \\
%             \emph{Present address:} of F. Author  %  if needed
           \and
           Paul-Christian Bürkner \at
              Excellence Cluster for Simulation Technology, University of Stuttgart, Germany \\
              \emph{Most of work was done while at Dept. of Computer Science at Aalto University}
           \and
           Michael R.\ Andersen \at
              Department of Applied Mathematics and Computer Science, Technical University of Denmark, Denmark
           \and
           Arno Solin \at
              Department of Computer Science, Aalto University, Finland
           \and
           Aki Vehtari \at
              Department of Computer Science, Aalto University, Finland\\
              \email{aki.vehtari@aalto.fi}
}

@@

The Mat\'ern class of isotropic covariance functions is given by, 
%
\begin{align*}
k_{\nu}(r)&=\alpha \, \frac{2^{1-\nu}}{\Gamma(\nu)}\left(\frac{\sqrt{2\nu}r}{\ell}\right)^{\!\nu} \! K_{\nu} \left(\frac{\sqrt{2\nu}r}{\ell}\right),
\end{align*}
%
where $\nu > 0$ is the order the kernel, $K_{\nu}$ the modified Bessel function of the second kind, and the $\ell > 0$ and $\alpha > 0$ are the length-scale and magnitude (marginal variance), respectively, of the kernel. The particular case where $\nu=\infty$, $\nu=3/2$ and $\nu=5/2$ are probably the most commonly used kernels \citep{rasmussen2006gaussian}, 
%
\begin{align*}
k_{\infty}(r)&=\alpha \exp \left( -\frac{1}{2} \frac{r^2}{\ell^2}\right),  \\
%k_{\frac{1}{2}}(r)&=\alpha \exp \left( -\frac{r}{\ell}\right),  \\
k_{\frac{3}{2}}(r)&=\alpha\left(1+\frac{\sqrt{3}r}{\ell}\right) \exp\! \left( -\frac{\sqrt{3}r}{\ell}\right),   \\
k_{\frac{5}{2}}(r)&=\alpha\left(1+\frac{\sqrt{5}r}{\ell} + \frac{5r^2}{3\ell^2}\right) \exp\! \left( -\frac{\sqrt{5}r}{\ell}\right). 
\end{align*}
%
The former is commonly known as the squared exponential or exponentiated quadratic covariance function. As an example, assuming the Euclidean distance between observations, $r=|\bm{x}-\bm{x}'|_{L2}=\sqrt{\sum_{i=1}^{D}(x_i-x_i')^2}$, the kernel $K_{\nu}$ written above takes the form
%
\begin{align*}
k_{\infty}(r) = \alpha & \exp \left( -\frac{1}{2} \sum_{i=1}^{D}\frac{(x_i-x_i')^2}{\ell_i^2}\right).
%k_{\frac{1}{2}}(r) = \alpha & \exp \left( -\sum_{i=1}^{D}\frac{(x_i-x_i')}{\ell_i}\right),\\
%k_{\frac{3}{2}}(r) = \alpha & \left(1+\sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}\;\right)\\
%& \times \exp\! \left( - \sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}\;\right),\\
%k_{\frac{5}{2}}(r) = \alpha & \left(1+\sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}+\frac{5}{3} \sum_{i=1}^{D}\frac{(x_i-x_i')^2}{\ell_i^2}\;\right)\\
%& \times \exp\! \left( - \sqrt{\sum_{i=1}^{D}\frac{5(x_i-x_i')^2}{\ell_i^2}}\;\right).
\end{align*}
%
Notice that the previous expressions $k_{\infty}(r)$ has been easily generalized to using a multidimensional length-scale $\bm{\ell}\in {\rm I\!R}^D$. Using individual length-scales for each dimension turns an isotropic covariance function into a non-isotropic covariance function. That is, for a non-isotropic covariance function, the smoothness may vary across different input dimensions.

Stationary covariance functions can be represented in terms of their spectral densities \citep[see, e.g.,][]{rasmussen2006gaussian}. In this sense, the covariance function of a stationary process can be represented as the Fourier transform of a positive finite measure \citep[\textit{Bochner's theorem}; see, e.g., ][]{akhiezer1993theory}. If this measure has a density, it is known as the spectral density of the covariance function, and the covariance function and the spectral density are Fourier duals, known as the \textit{Wiener-Khintchine theorem} \citep{rasmussen2006gaussian}. The spectral density functions associated with the Mat\'ern class of covariance functions are given by
%
\begin{align*}
S_{\nu}(\bm{\omega})&= \alpha \, \frac{2^D\pi^{D/2}\Gamma(\nu+D/2)(2\nu)^{\nu}}{\Gamma(\nu)\, \ell^{2\nu}}\left(\frac{2\nu}{\ell^2}+4\pi^2\bm{\omega}^\intercal \bm{\omega} \right)^{-(\nu+D/2)}
\end{align*}
%
in $D$ dimensions, where vector $\bm{\omega}\in {\rm I\!R^D}$ is in the frequency domain, and $\ell$ and $\alpha$ are the length-scale and magnitude (marginal variance), respectively, of the kernel. The particular cases, where $\nu=\infty$, $\nu=1/2$, $\nu=3/2$ and $\nu=5/2$, take the form
%
\begin{align}
S_{\infty}(\bm{\omega})&= \alpha\,(\sqrt{2\pi})^D  \ell^D \exp(-\frac{1}{2}\ell^2 \bm{\omega}^\intercal \bm{\omega}), \label{eq_specdens_inf}  \\
%S_{\frac{1}{2}}(\bm{\omega})&= \alpha\,\frac{2^D\pi^{D/2}\Gamma(\frac{D+1}{2})}{\sqrt{\pi}\ell}\left(\frac{1}{\ell^2}+\bm{\omega}^\intercal \bm{\omega} \right)^{-\frac{D+1}{2}} \label{eq_specdens_12},     \\
S_{\frac{3}{2}}(\bm{\omega})&= \alpha\,\frac{2^D\pi^{D/2}\Gamma(\frac{D+3}{2}){3}^{3/2}}{\frac{1}{2}\sqrt{\pi}\ell^3}\left(\frac{3}{\ell^2}+\bm{\omega}^\intercal \bm{\omega} \right)^{-\frac{D+3}{2}} \label{eq_specdens_32},     \\
S_{\frac{5}{2}}(\bm{\omega})&= \alpha\,\frac{2^D\pi^{D/2}\Gamma(\frac{D+5}{2}){5}^{5/2}}{\frac{3}{4}\sqrt{\pi}\ell^5}\left(\frac{5}{\ell^2}+\bm{\omega}^\intercal \bm{\omega} \right)^{-\frac{D+5}{2}}. \label{eq_specdens_52}   
\end{align}
%
For instance, with input dimensionality $D=3$ and $\bm{\omega}=(\omega_1,\omega_2,\omega_3)^\intercal$, the spectral densities written above take the form
%
\begin{align*}
S_{\infty}(\bm{\omega})&= \alpha \, (2\pi)^{3/2}  \prod_{i=1}^{3} \ell_i  \exp\!\left(-\frac{1}{2} \sum_{i=1}^{3} \ell_i^2 \omega_i^2 \right),   \\
%S_{\frac{1}{2}}(\bm{\omega})&= \alpha \, 8\,\pi \prod_{i=1}^{3}\ell_i\left(1+\sum_{i=1}^{3}\ell_i^2 \omega_i^2 \right)^{-2},    \\
S_{\frac{3}{2}}(\bm{\omega})&= \alpha \, 32\,\pi {3}^{3/2}\prod_{i=1}^{3}\ell_i\left(3+\sum_{i=1}^{3}\ell_i^2 \omega_i^2 \right)^{-3},   \\
S_{\frac{5}{2}}(\bm{\omega})&= \alpha \, \frac{64}{3}\,\pi {5}^{5/2}\prod_{i=1}^{3}\ell_i\left(5+\sum_{i=1}^{3}\ell_i^2 \omega_i^2 \right)^{-4}.
\end{align*}
%
where individual length-scales $\ell_i$ for each frequency dimension $\omega_i$ have been used.

@@

Within $\Omega$, we can write any stationary covariance function with input values $x,x' \in \Omega$ as
%
\begin{equation}\label{eq_approxcov}
k(x,x') = \sum_{j=1}^\infty S_{\theta}(\sqrt{\lambda_j}) \phi_j(x) \phi_j(x'),
\end{equation} 
%
where $S_{\theta}$ is the spectral density of the stationary covariance function $k$ (see Section \ref{ch4_sec_cov}) and $\theta$ is the set of hyperparameters of $k$ \citep{rasmussen2006gaussian}. The terms $\{\lambda_j\}_{j=1}^{\infty}$ and $\{\phi_j(x)\}_{j=1}^{\infty}$ are the sets of eigenvalues and eigenvectors, respectively, of the Laplacian operator in the given domain $\Omega$. Namely, they satisfy the following eigenvalue problem in $\Omega$ when applying the Dirichlet boundary condition (other boundary conditions could be used as well)
%
\begin{align}\label{eq_eigenproblem}
\begin{split}
-\nabla^2 \phi_j(x)&=\lambda_j \phi_j(x), \hspace{1cm}  x \in \Omega \\ 
\phi_j(x)&= 0, \hspace{1.85cm}   x \notin \Omega.
\end{split}
\end{align} 
%
The eigenvalues $\lambda_j>0$ are real and positive because the Laplacian is a positive definite Hermitian operator, and the eigenfunctions $\phi_j$ for the eigenvalue problem in eq.~\eqref{eq_eigenproblem} are sinusoidal functions. The solution to the eigenvalue problem is independent of  the specific choice of covariance function and is given by
%
\begin{align}
\lambda_j&=\left(\frac{j\pi}{2L}\right)^{\!2}, \label{eq_eigenvalue}\\
\phi_j(x)&=\sqrt{\frac{1}{L}}\, \sin\!\!\left(\sqrt{\lambda_j}(x+L)\right). \label{eq_eigenfunction}
\end{align}

If we truncate the sum in eq.~\eqref{eq_approxcov} to the first $m$ terms, the approximate covariance function becomes
%
\begin{equation}
k(x,x') \approx \sum_{j=1}^m S_{\theta}(\sqrt{\lambda_j}) \phi_j(x) \phi_j(x') = \bm{\phi}(x)^\intercal \Delta \bm{\phi}(x'), \nonumber
\end{equation}
%
where $\bm{\phi}(x)=\{\phi_j(x)\}_{j=1}^{m} \in {\rm I\!R}^{m}$ is the column vector of basis functions, and $\Delta  \in {\rm I\!R}^{m\times m}$ is a diagonal matrix of the spectral density evaluated at the square root of the eigenvalues, that is, $S_{\theta}(\sqrt{\lambda_j})$,
%
\begin{align}
\Delta =  \begin{bmatrix}
    S_{\theta}(\sqrt{\lambda_1}) & & \\
    & \ddots & \nonumber \\
    & & S_{\theta}(\sqrt{\lambda_m}) \\
  \end{bmatrix}.
\end{align}

Thus, the Gram matrix $\bm{K}$ for the covariance function $k$ for a set of observations $i=1,\ldots,n$ and corresponding input values $\{x_i\}_{i=1}^{n}$ can be represented as
%
\begin{equation}
 \bm{K} = \Phi \Delta \Phi^\intercal, \nonumber
\end{equation}
%
where $\Phi \in {\rm I\!R}^{n\times m}$ is the matrix of eigenfunctions $\phi_j(x_i)$
%
\begin{align}
\Phi =  \left[ {\begin{array}{ccc}
   \phi_1(x_1) & \cdots & \phi_m(x_1)  \\
    \vdots &\ddots & \vdots  \nonumber \\ 
    \phi_1(x_n) & \cdots & \phi_m(x_n) \\
  \end{array} } \right].
\end{align}
As a result, the model for $f$ can be written as
%
\begin{equation}
\bm{f} \sim \Normal(\bm{\mu},\Phi \Delta \Phi^\intercal). \nonumber
\end{equation}
This equivalently leads to a linear representation of $f$ via
%
\begin{equation}\label{eq_approxf}
f(x) \approx \sum_{j=1}^m \left( S_{\theta}(\sqrt{\lambda_j})\right)^{\frac{1}{2}} \phi_j(x) \beta_j,
\end{equation}
%
where $\beta_j \sim \Normal(0,1)$. Thus, the function $f$ is approximated with a finite basis function expansion (using the eigenfunctions $\phi_j$ of the Laplace operator), scaled by the square root of spectral density values. A key property of this approximation is that the eigenfunctions $\phi_j$ do not depend on the hyperparameters of the covariance function $\theta$. Instead, the only dependence of the model on $\theta$ is through the spectral density $S_{\theta}$. The eigenvalues $\lambda_j$ are monotonically increasing with $j$ and $S_{\theta}$ goes rapidly to zero for bounded covariance functions. Therefore, eq.~\eqref{eq_approxf} can be expected to be a good approximation for a finite number of $m$ terms in the series as long as the inputs values $x_i$ are not too close to the boundaries $-L$ and $L$ of $\Omega$. The computational cost of evaluating the log posterior density of univariate HSGPs scales as $O(nm + m)$, where $n$ is the number of observations and $m$ the number of basis functions.

@@

The results from the previous section can be generalized to a multidimensional input space with compact support, $\Omega=[-L_1,L_1] \times \dots \times [-L_D,L_D]$ and Dirichlet boundary conditions. 
In a $D$-dimensional input space, the total number of eigenfunctions and eigenvalues in the approximation is equal to the number of $D$-tuples, that is, possible combinations of univariate eigenfunctions over all dimensions. The number of $D$-tuples is given by 
%
\begin{align} \label{eq_m_multi}
m^{\ast} = \prod_{d=1}^{D} m_d,
\end{align}
%
where $m_d$ is the number of basis function for the dimension $d$. Let $\mathbb{S}\in {\rm I\!N}^{m^{\ast} \times D}$ be the matrix of all those $D$-tuples. For example, suppose we have $D=3$ dimensions and use $m_{1}=2$, $m_{2}=2$ and $m_{3}=3$ eigenfunctions and eigenvalues for the first, second and third dimension, respectively. Then, the number of multivariate eigenfunctions and eigenvalues is $m^{\ast} = m_{1} \cdot m_{2} \cdot m_{3} = 12$ and the matrix $\mathbb{S}\in {\rm I\!N}^{12 \times 3}$ is given by
%
%\begin{align}\small
%\mathbb{S}=
%\left[ {\begin{array}{ccc}
%1 & 1 & 1 \nonumber \\
%1 & 1 & 2 \\
%1 & 1 & 3 \\
%1 & 2 & 1 \\
%1 & 2 & 2 \\
%1 & 2 & 3 \\
%2 & 1 & 1 \\
%2 & 1 & 2 \\
%2 & 1 & 3 \\
%2 & 2 & 1 \\
%2 & 2 & 2 \\
%2 & 2 & 3 
%\end{array} } \right].
%\end{align} 
%
\begin{align}\small
\mathbb{S}=
\left[ {\begin{array}{cccccccccccc}
1 & 1 & 1 & 1 & 1 & 1 & 2 & 2 & 2 & 2 & 2 & 2 \\
1 & 1 & 1 & 2 & 2 & 2 & 1 & 1 & 1 & 2 & 2 & 2 \\
1 & 2 & 3 & 1 & 2 & 3 & 1 & 2 & 3 & 1 & 2 & 3
\end{array} } \right]^\intercal.
\end{align}

>>>>

\institute{Gabriel Riutort-Mayol \at
              Foundation for the Promotion of Health and Biomedical Research of Valencia Region (FISABIO), Spain\\
              \emph{Corresponding Author} \\
              \email{riutort\_gab@gva.es; gabriuma@gmail.com}       %  \\
%             \emph{Present address:} of F. Author  %  if needed
           \and
           Paul-Christian Bürkner \at
              Excellence Cluster for Simulation Technology, University of Stuttgart, Germany \\
              \emph{Most of work was done while at Dept. of Computer Science at Aalto University}
           \and
           Michael R.\ Andersen \at
              Department of Applied Mathematics and Computer Science, Technical University of Denmark, Denmark
           \and
           Arno Solin \at
              Department of Computer Science, Aalto University, Finland
           \and
           Aki Vehtari \at
              Department of Computer Science, Aalto University, Finland\\
              \email{aki.vehtari@aalto.fi}
}

@@

The Mat\'ern class of isotropic covariance functions is given by, 
%
\begin{align*}
k_{\nu}(r)&=\alpha \, \frac{2^{1-\nu}}{\Gamma(\nu)}\left(\frac{\sqrt{2\nu}r}{\ell}\right)^{\!\nu} \! K_{\nu} \left(\frac{\sqrt{2\nu}r}{\ell}\right),
\end{align*}
%
where $\nu > 0$ is the order the kernel, $K_{\nu}$ the modified Bessel function of the second kind, and the $\ell > 0$ and $\alpha > 0$ are the length-scale and magnitude (marginal variance), respectively, of the kernel. The particular case where $\nu=\infty$, $\nu=3/2$ and $\nu=5/2$ are probably the most commonly used kernels \citep{rasmussen2006gaussian}, 
%
\begin{align*}
k_{\infty}(r)&=\alpha \exp \left( -\frac{1}{2} \frac{r^2}{\ell^2}\right),  \\
%k_{\frac{1}{2}}(r)&=\alpha \exp \left( -\frac{r}{\ell}\right),  \\
k_{\frac{3}{2}}(r)&=\alpha\left(1+\frac{\sqrt{3}r}{\ell}\right) \exp\! \left( -\frac{\sqrt{3}r}{\ell}\right),   \\
k_{\frac{5}{2}}(r)&=\alpha\left(1+\frac{\sqrt{5}r}{\ell} + \frac{5r^2}{3\ell^2}\right) \exp\! \left( -\frac{\sqrt{5}r}{\ell}\right). 
\end{align*}
%
The former is commonly known as the squared exponential or exponentiated quadratic covariance function. As an example, assuming the Euclidean distance between observations, $r=|\bm{x}-\bm{x}'|_{L2}=\sqrt{\sum_{i=1}^{D}(x_i-x_i')^2}$, the kernel $k_{\infty}$ written above takes the form
%
\begin{align*}
k_{\infty}(r) = \alpha & \exp \left( -\frac{1}{2} \sum_{i=1}^{D}\frac{(x_i-x_i')^2}{\ell_i^2}\right).
%k_{\frac{1}{2}}(r) = \alpha & \exp \left( -\sum_{i=1}^{D}\frac{(x_i-x_i')}{\ell_i}\right),\\
%k_{\frac{3}{2}}(r) = \alpha & \left(1+\sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}\;\right)\\
%& \times \exp\! \left( - \sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}\;\right),\\
%k_{\frac{5}{2}}(r) = \alpha & \left(1+\sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}+\frac{5}{3} \sum_{i=1}^{D}\frac{(x_i-x_i')^2}{\ell_i^2}\;\right)\\
%& \times \exp\! \left( - \sqrt{\sum_{i=1}^{D}\frac{5(x_i-x_i')^2}{\ell_i^2}}\;\right).
\end{align*}
%
Notice that the previous expressions $k_{\infty}(r)$ has been easily generalized to using a multidimensional length-scale $\bm{\ell}\in {\rm I\!R}^D$. Using individual length-scales for each dimension turns an isotropic covariance function into a non-isotropic covariance function. That is, for a non-isotropic covariance function, the smoothness may vary across different input dimensions.

Stationary covariance functions can be represented in terms of their spectral densities \citep[see, e.g.,][]{rasmussen2006gaussian}. In this sense, the covariance function of a stationary process can be represented as the Fourier transform of a positive finite measure \citep[\textit{Bochner's theorem}; see, e.g., ][]{akhiezer1993theory}. If this measure has a density, it is known as the spectral density of the covariance function, and the covariance function and the spectral density are Fourier duals, known as the \textit{Wiener-Khintchine theorem} \citep{rasmussen2006gaussian}. The spectral density functions associated with the Mat\'ern class of covariance functions are given by
%
\begin{align*}
s_{\nu}(\bm{\omega})&= \alpha \, \frac{2^D\pi^{D/2}\Gamma(\nu+D/2)(2\nu)^{\nu}}{\Gamma(\nu)\, \ell^{2\nu}}\left(\frac{2\nu}{\ell^2}+4\pi^2\bm{\omega}^\intercal \bm{\omega} \right)^{-(\nu+D/2)}
\end{align*}
%
in $D$ dimensions, where vector $\bm{\omega}\in {\rm I\!R^D}$ is in the frequency domain, and $\ell$ and $\alpha$ are the length-scale and magnitude (marginal variance), respectively, of the kernel. The particular cases, where $\nu=\infty$, $\nu=1/2$, $\nu=3/2$ and $\nu=5/2$, take the form
%
\begin{align}
s_{\infty}(\bm{\omega})&= \alpha\,(\sqrt{2\pi})^D  \ell^D \exp(-\frac{1}{2}\ell^2 \bm{\omega}^\intercal \bm{\omega}), \label{eq_specdens_inf}  \\
%s_{\frac{1}{2}}(\bm{\omega})&= \alpha\,\frac{2^D\pi^{D/2}\Gamma(\frac{D+1}{2})}{\sqrt{\pi}\ell}\left(\frac{1}{\ell^2}+\bm{\omega}^\intercal \bm{\omega} \right)^{-\frac{D+1}{2}} \label{eq_specdens_12},     \\
s_{\frac{3}{2}}(\bm{\omega})&= \alpha\,\frac{2^D\pi^{D/2}\Gamma(\frac{D+3}{2}){3}^{3/2}}{\frac{1}{2}\sqrt{\pi}\ell^3}\left(\frac{3}{\ell^2}+\bm{\omega}^\intercal \bm{\omega} \right)^{-\frac{D+3}{2}} \label{eq_specdens_32},     \\
s_{\frac{5}{2}}(\bm{\omega})&= \alpha\,\frac{2^D\pi^{D/2}\Gamma(\frac{D+5}{2}){5}^{5/2}}{\frac{3}{4}\sqrt{\pi}\ell^5}\left(\frac{5}{\ell^2}+\bm{\omega}^\intercal \bm{\omega} \right)^{-\frac{D+5}{2}}. \label{eq_specdens_52}   
\end{align}
%
For instance, with input dimensionality $D=3$ and $\bm{\omega}=(\omega_1,\omega_2,\omega_3)^\intercal$, the spectral densities written above take the form
%
\begin{align*}
s_{\infty}(\bm{\omega})&= \alpha \, (2\pi)^{3/2}  \prod_{i=1}^{3} \ell_i  \exp\!\left(-\frac{1}{2} \sum_{i=1}^{3} \ell_i^2 \omega_i^2 \right),   \\
%s_{\frac{1}{2}}(\bm{\omega})&= \alpha \, 8\,\pi \prod_{i=1}^{3}\ell_i\left(1+\sum_{i=1}^{3}\ell_i^2 \omega_i^2 \right)^{-2},    \\
s_{\frac{3}{2}}(\bm{\omega})&= \alpha \, 32\,\pi {3}^{3/2}\prod_{i=1}^{3}\ell_i\left(3+\sum_{i=1}^{3}\ell_i^2 \omega_i^2 \right)^{-3},   \\
s_{\frac{5}{2}}(\bm{\omega})&= \alpha \, \frac{64}{3}\,\pi {5}^{5/2}\prod_{i=1}^{3}\ell_i\left(5+\sum_{i=1}^{3}\ell_i^2 \omega_i^2 \right)^{-4}.
\end{align*}
%
where individual length-scales $\ell_i$ for each frequency dimension $\omega_i$ have been used.

@@

Within $\Omega$, we can write any stationary covariance function with input values $x,x' \in \Omega$ as
%
\begin{equation}\label{eq_approxcov}
k(x,x') = \sum_{j=1}^\infty s_{\theta}(\sqrt{\lambda_j}) \phi_j(x) \phi_j(x'),
\end{equation} 
%
where $s_{\theta}$ is the spectral density of the stationary covariance function $k$ (see Section \ref{ch4_sec_cov}) and $\theta$ is the set of hyperparameters of $k$ \citep{rasmussen2006gaussian}. The terms $\{\lambda_j\}_{j=1}^{\infty}$ and $\{\phi_j(x)\}_{j=1}^{\infty}$ are the sets of eigenvalues and eigenvectors, respectively, of the Laplacian operator in the given domain $\Omega$. Namely, they satisfy the following eigenvalue problem in $\Omega$ when applying the Dirichlet boundary condition (other boundary conditions could be used as well)
%
\begin{align}\label{eq_eigenproblem}
\begin{split}
-\nabla^2 \phi_j(x)&=\lambda_j \phi_j(x), \hspace{1cm}  x \in \Omega \\ 
\phi_j(x)&= 0, \hspace{1.85cm}   x \notin \Omega.
\end{split}
\end{align} 
%
The eigenvalues $\lambda_j>0$ are real and positive because the Laplacian is a positive definite Hermitian operator, and the eigenfunctions $\phi_j$ for the eigenvalue problem in eq.~\eqref{eq_eigenproblem} are sinusoidal functions. The solution to the eigenvalue problem is independent of  the specific choice of covariance function and is given by
%
\begin{align}
\lambda_j&=\left(\frac{j\pi}{2L}\right)^{\!2}, \label{eq_eigenvalue}\\
\phi_j(x)&=\sqrt{\frac{1}{L}}\, \sin\!\!\left(\sqrt{\lambda_j}(x+L)\right). \label{eq_eigenfunction}
\end{align}

If we truncate the sum in eq.~\eqref{eq_approxcov} to the first $m$ terms, the approximate covariance function becomes
%
\begin{equation}
k(x,x') \approx \sum_{j=1}^m s_{\theta}(\sqrt{\lambda_j}) \phi_j(x) \phi_j(x') = \bm{\phi}(x)^\intercal \Delta \bm{\phi}(x'), \nonumber
\end{equation}
%
where $\bm{\phi}(x)=\{\phi_j(x)\}_{j=1}^{m} \in {\rm I\!R}^{m}$ is the column vector of basis functions, and $\Delta  \in {\rm I\!R}^{m\times m}$ is a diagonal matrix of the spectral density evaluated at the square root of the eigenvalues, that is, $s_{\theta}(\sqrt{\lambda_j})$,
%
\begin{align}
\Delta =  \begin{bmatrix}
    s_{\theta}(\sqrt{\lambda_1}) & & \\
    & \ddots & \nonumber \\
    & & s_{\theta}(\sqrt{\lambda_m}) \\
  \end{bmatrix}.
\end{align}

Thus, the Gram matrix $\bm{K}$ for the covariance function $k$ for a set of observations $i=1,\ldots,n$ and corresponding input values $\{x_i\}_{i=1}^{n}$ can be represented as
%
\begin{equation}
 \bm{K} = \Phi \Delta \Phi^\intercal, \nonumber
\end{equation}
%
where $\Phi \in {\rm I\!R}^{n\times m}$ is the matrix of eigenfunctions $\phi_j(x_i)$
%
\begin{align}
\Phi =  \left[ {\begin{array}{ccc}
   \phi_1(x_1) & \cdots & \phi_m(x_1)  \\
    \vdots &\ddots & \vdots  \nonumber \\ 
    \phi_1(x_n) & \cdots & \phi_m(x_n) \\
  \end{array} } \right].
\end{align}
As a result, the model for $f$ can be written as
%
\begin{equation}
\bm{f} \sim \Normal(\bm{\mu},\Phi \Delta \Phi^\intercal). \nonumber
\end{equation}
This equivalently leads to a linear representation of $f$ via
%
\begin{equation}\label{eq_approxf}
f(x) \approx \sum_{j=1}^m \left( s_{\theta}(\sqrt{\lambda_j})\right)^{\frac{1}{2}} \phi_j(x) \beta_j,
\end{equation}
%
where $\beta_j \sim \Normal(0,1)$. Thus, the function $f$ is approximated with a finite basis function expansion (using the eigenfunctions $\phi_j$ of the Laplace operator), scaled by the square root of spectral density values. A key property of this approximation is that the eigenfunctions $\phi_j$ do not depend on the hyperparameters of the covariance function $\theta$. Instead, the only dependence of the model on $\theta$ is through the spectral density $s_{\theta}$. The eigenvalues $\lambda_j$ are monotonically increasing with $j$ and $s_{\theta}$ goes rapidly to zero for bounded covariance functions. Therefore, eq.~\eqref{eq_approxf} can be expected to be a good approximation for a finite number of $m$ terms in the series as long as the inputs values $x_i$ are not too close to the boundaries $-L$ and $L$ of $\Omega$. The computational cost of evaluating the log posterior density of univariate HSGPs scales as $O(nm + m)$, where $n$ is the number of observations and $m$ the number of basis functions.

@@

The results from the previous section can be generalized to a multidimensional input space with compact support, $\Omega=[-L_1,L_1] \times \dots \times [-L_D,L_D]$ and Dirichlet boundary conditions. 
In a $D$-dimensional input space, the total number of eigenfunctions and eigenvalues in the approximation is equal to the number of $D$-tuples, that is, possible combinations of univariate eigenfunctions over all dimensions. The number of $D$-tuples is given by 
%
\begin{align} \label{eq_m_multi}
m^{\ast} = \prod_{d=1}^{D} m_d,
\end{align}
%
where $m_d$ is the number of basis function for the dimension $d$. Let $\mathbb{S}\in {\rm I\!N}^{m^{\ast} \times D}$ be the matrix of all those $D$-tuples. For example, suppose we have $D=3$ dimensions and use $m_{1}=2$, $m_{2}=2$ and $m_{3}=3$ eigenfunctions and eigenvalues for the first, second and third dimension, respectively. Then, the number of multivariate eigenfunctions and eigenvalues is $m^{\ast} = m_{1} \cdot m_{2} \cdot m_{3} = 12$ and the matrix $\mathbb{S}\in {\rm I\!N}^{12 \times 3}$ is given by
%
\begin{align}\small
\mathbb{S}=
\left[ {\begin{array}{cccccccccccc}
1 & 1 & 1 & 1 & 1 & 1 & 2 & 2 & 2 & 2 & 2 & 2 \\
1 & 1 & 1 & 2 & 2 & 2 & 1 & 1 & 1 & 2 & 2 & 2 \\
1 & 2 & 3 & 1 & 2 & 3 & 1 & 2 & 3 & 1 & 2 & 3
\end{array} } \right]^\intercal.
\end{align}

Each multivariate eigenfunction $\phi^{\ast}_j:\Omega \rightarrow {\rm I\!R}$ corresponds to the product of the univariate eigenfunctions whose indices corresponds to the elements of the $D$-tuple $\mathbb{S}_{j\cdotp}$, and each multivariate eigenvalue $\bm{\lambda}^{\ast}_j$ is a $D$-vector with elements that are the univariate eigenvalues whose indices correspond to the elements of the $D$-tuple $\mathbb{S}_{j\bm{\cdotp}}$. Thus, for $\bm{x}=\{x_d\}_{d=1}^D \in \Omega$ and $j=1,2,\ldots,m^{\ast}$, we have 
%
\begin{align}
\bm{\lambda}^{\ast}_j &= \left\{ \lambda_{\mathbb{S}_{jd}} \right\}_{d=1}^D =  \left\{ \left(\frac{\pi \mathbb{S}_{jd}}{2L_d}\right)^{\!2} \right\}_{d=1}^D, \label{eq_eigenvalue_multi} \\
%
\phi^{\ast}_j(\bm{x}) &= \prod_{d=1}^{D} \phi_{\mathbb{S}_{jd}}(x_d) = \prod_{d=1}^{D} \sqrt{\frac{1}{L_d}} \sin\!\left(\sqrt{\lambda_{\mathbb{S}_{jd}}}(x_d+L_d)\right). \label{eq_eigenfunction_multi}
\end{align}
%
The approximate covariance function is then represented as
%
\begin{equation}\label{eq_approxcov_multi}
k(\bm{x},\bm{x}') \approx \sum_{j=1}^{m^{\ast}} 
s^{\ast}_{\theta}\left(\sqrt{\bm{\lambda}^{\ast}_j}\right)
\phi^{\ast}_j(\bm{x}) \phi^{\ast}_j(\bm{x}'),
\end{equation}
%
where $s^{\ast}_{\theta}$ is the spectral density of the $D$-dimensional covariance function (see Section \ref{ch4_sec_cov}) as a function of $\sqrt{\bm{\lambda}^{\ast}_j}$ that denotes the element-wise square root of the vector $\bm{\lambda}^{\ast}_j$. We can now write the approximate series expansion of the multivariate function $f$ as
%
\begin{equation}\label{eq_approxf_multi}
f(\bm{x}) \approx \sum_{j=1}^{m^{\ast}} 
\left( s^{\ast}_{\theta} \left(\sqrt{\bm{\lambda}^{\ast}_j} \right)\right)^{\! \frac{1}{2}} \phi^{\ast}_j(\bm{x}) \beta_j, 
\end{equation}
%
where, again, $\beta_j \sim \Normal(0,1)$. The computational cost of evaluating the log posterior density of multivariate HSGPs scales as $O(n m^{\ast} + m^{\ast})$, where $n$ is the number of observations and $m^{\ast}$ is the number of multivariate basis functions. Although this still implies linear scaling in $n$, the approximation is more costly than in the univariate case, as $m^{\ast}$ is the product of the number of univariate basis functions over the input dimensions and grows exponentially with respect to the number of  dimensions.
